% Three-page project write-up, "Reliable ML with Unreliable Black Boxes".
% Core = the bucket (grid-mixture) streaming certificate and its provable
% guarantees (Lemma 1 + proof), plus three preliminary findings from the talk.
% Everything else is pushed to the appendices. Build: latexmk -pdf writeup.tex
%
% Numbers are NEVER hand-typed: every inline scalar resolves through a
% script-generated macro (\input paper_macros.tex / grid_certificate macros).
\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\newtheorem{lemma}{Lemma}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{parskip}
\usepackage[round]{natbib}
\usepackage[hidelinks]{hyperref}

\graphicspath{%
  {../results/figures/}%
  {../results/counter_decode/self_calibration/}%
}
\newcommand{\tabdir}{../results/tables}
\input{../results/tables/paper_macros.tex}
\input{../data/grid_certificate/macros.tex}

\title{Reading the Decoder: A Distribution-Free Streaming Certificate\\
for an LLM's Effective Decoding Temperature}
\author{Matthew Khoriaty}
\date{Project write-up, \emph{Reliable ML with Unreliable Black Boxes}}

\begin{document}
\maketitle

\begin{abstract}
An autoregressive language model is trained to \emph{predict} text, so when it
\emph{generates}, it has an incentive to model the decoder that turns its logits
into tokens (a temperature, a top-$k$ cutoff, a banned-token list). This
write-up reports preliminary steps toward reading that decoder off the model's
own logits, training-free. Our central result is a \textbf{partial proof}: a
streaming estimator built from a grid (``bucket'') of fixed temperature experts
is exactly Vovk's Aggregating Algorithm for log loss, and therefore carries two
\emph{per-sequence, distribution-free} guarantees on arbitrary token streams ---
a forecast-regret bound of $\ln G$ against the best bucket in hindsight, and a
convex level-set error bar around the grid MLE that needs no distributional
assumptions (Lemma~\ref{lem:certificate}). We then report three preliminary
findings, framed as questions: \textbf{(1)}~the model tracks a token blacklist
and a top-$k$ cutoff from its own constrained text; \textbf{(2)}~a streaming
read-out of $\hat\beta_t$ predicts known-temperature self-text near-optimally but
finds little to recalibrate on natural text, where Qwen-2.5-3B is
temperature-robust; and \textbf{(3)}~closed-loop correction self-neutralizes, so
using the read-out to \emph{generate} better is still open. All numbers are
script-generated and cited by name.
\end{abstract}

% =====================================================================
\section{The research question}
% =====================================================================
At each step an autoregressive language model emits a distribution over next
tokens; \emph{decoding} chooses how to sample from it. Because models are trained
to predict text rather than to generate it, a model that reads back its own
generated prefix has an incentive to infer the decoder that produced it and
adjust its predictions accordingly \citep{basu2021mirostat}. The question this
project addresses is whether that decoder --- concretely, an effective
inverse-temperature $\beta = 1/T$ --- can be \emph{read off the model's own
logits online, training-free, and with guarantees}, and whether the read-out is
useful for prediction or generation. This sits at the intersection of post-hoc
calibration \citep{guo2017calibration,liu2024temperature} and the
algorithms-with-predictions viewpoint \citep{mitzenmacher2020algorithms} of the
course, in which a possibly-unreliable prediction augments a downstream decision;
what we then most want is a \emph{worst-case, per-sequence} guarantee that holds
even when the stream is misspecified. Section~\ref{sec:cert} delivers one;
Section~\ref{sec:findings} reports three preliminary experiments, with supporting
machinery (a fast point filter, a corrector, control experiments) in the
appendices.

% =====================================================================
\section{The bucket streaming certificate}\label{sec:cert}
% =====================================================================
Fix a grid (a set of ``buckets'') of inverse-temperatures
$\beta_1 < \dots < \beta_G$, log-spaced over a bounded range. Treat each
$\beta_g$ as a fixed \emph{expert} that forecasts the next token with probability
$P_g(x_t) = \mathrm{softmax}(\beta_g\,\ell_t)[x_t]$ on the model's \emph{raw}
logits $\ell_t \in \mathbb{R}^V$ (no rescaling --- fixed experts are what make
the guarantees hold). Let
\[
  L_n(\beta) = \sum_{t\le n} \big[\mathrm{lse}(\beta\ell_t) - \beta\,\ell_t[x_t]\big],
  \qquad \mathrm{lse}(\cdot) = \log\textstyle\sum_v \exp(\cdot),
\]
be the cumulative log loss of bucket $\beta$, let $\pi_0$ be a prior on the grid,
and let the streaming forecaster predict $x_t$ with probability
$\sum_g w_g(t)\,P_g(x_t)$, where $w_g(t) \propto \pi_0[g]\,e^{-L_{t-1}(\beta_g)}$
is the Bayes posterior over buckets. This is precisely Vovk's Aggregating
Algorithm for log loss \citep{vovk1998game,cesabianchi2006prediction}, and the
following holds for \emph{every} logit/token sequence --- with no assumption that
tokens were sampled from any $P_\beta$, or from anything at all.

\begin{lemma}[per-sequence certificate]\label{lem:certificate}
For any logit sequence $\ell_{1:n}$ and any token sequence $x_{1:n}$:
(i)~the forecaster's cumulative log loss is
$-\ln \sum_g \pi_0[g]\, e^{-L_n(\beta_g)} \le L_n(\beta_g) + \ln(1/\pi_0[g])$
for every $g$; with uniform $\pi_0$ the regret against the best bucket in
hindsight is at most $\ln G$.
(ii)~With uniform $\pi_0$ the posterior mode equals $\arg\min_g L_n(\beta_g)$,
the best-fit (\emph{effective}) inverse temperature on the grid, exactly.
(iii)~$L_n$ is convex with $L_n''(\beta) = \sum_t \mathrm{Var}_{q_{\beta,t}}[\ell_t]$,
$q_{\beta,t} = \mathrm{softmax}(\beta\ell_t)$, so for any $c>0$ the sub-level set
$\{\beta : L_n(\beta) \le \min L_n + c\}$ is an interval containing the sequence
MLE; its endpoints, read off the stored grid loss profile, are a data-dependent
error bar that needs no distributional assumptions.
\end{lemma}

\begin{proof}
(i)~The forecaster's per-step log probabilities telescope:
\[
  \textstyle\sum_t \ln \sum_g w_g(t) P_g(x_t)
  = \ln \sum_g \pi_0[g]\, e^{-L_n(\beta_g)}
  \ge \ln\big(\pi_0[g]\, e^{-L_n(\beta_g)}\big)\ \text{for each } g.
\]
(ii)~Immediate from $w_g(n) \propto e^{-L_n(\beta_g)}$.
(iii)~Each summand of $L_n$ is an $\mathrm{lse}$ of an affine function of $\beta$
minus an affine function, hence convex; differentiate twice.
\end{proof}

``Distribution-free'' here does double duty. Over the data process, the lemma
assumes nothing about how $x_{1:n}$ was produced. Over the estimator's own
assumptions, with uniform $\pi_0$ the mode in (ii) \emph{is} the grid MLE and the
bar in (iii) is an exact profile-likelihood interval (not a Gaussian
approximation); the prior survives only as the constant $c=\ln G$. We report the
mode, not the posterior mean: by (ii) it equals the best-fit bucket with zero
slack, whereas the mean of the (unimodal but skewed --- $e^{-L_n}$ is log-concave
by (iii)) posterior over a log-spaced grid is mildly biased.

\paragraph{Three honest caveats.}
First, the certificate is relative to the in-hindsight best-fit $\beta$, not the
dial a counterparty actually set; under a misspecified decoder (a top-$k$ cutoff,
a blacklist) no single ``true'' $\beta$ describes the stream, and the best-fit
value is exactly what ``effective temperature'' should mean there. The bar then
certifies \emph{fit to the observed stream}, and a narrow bar means $\beta$ is
\emph{identifiable}, not that any temperature fits well. Second, the interval in
(iii) widens exactly when peaked logits make $\beta$ unidentifiable
($\mathrm{Var}_{q}[\ell_t]\to 0$), so forecasting regret and estimation precision
degrade in opposite directions. Third, the guarantee attaches to the mixture, not
a point estimator (proper online predictors face an exp-concavity obstruction
that improper mixtures avoid \citep{foster2018logistic}), so transferring it to
the fast filter (Appendix~\ref{app:laplace}) is an empirical, pre-registered
claim.

\paragraph{Pre-registered validation.}
We committed four hypotheses, with sample sizes and thresholds, \emph{before}
generating any data (\texttt{paper/grid\_certificate\_prereg.md}); the bar uses
$c=\ln G$ on a $121$-bucket grid spanning $\log\beta\in[-3,3]$.
Table~\ref{tab:grid_certificate} reports the outcome. The bar covered the
continuum MLE on every well-specified synthetic stream (\gridCertCoverage); on
real Qwen-2.5-3B streams at $\beta{=}1$ the median width was
\gridCertMedianWidthBtOne{} in $\log\beta$ (the grid-resolution floor --- the bar
is far from vacuous where logits carry temperature information) and widened on
peaked-logit streams exactly as (iii) predicts. The transfer check to the fast
filter passed (\gridCertFaithfulAgree{} of estimates inside the bar), while the
original $\beta{=}1$-linearized filter landed inside on only
\gridCertOriginalAgree{} --- so the certificate extends to the corrected filter's
outputs, not the original's.

\begin{table}[t]
\centering
\input{\tabdir/tab_grid_certificate.tex}
\caption{Pre-registered evaluation of the certificate (hypotheses, sample sizes,
and thresholds committed before any data were generated). Real streams are
Qwen-2.5-3B, $32$ prompts, $1000$ tokens.}
\label{tab:grid_certificate}
\end{table}

% =====================================================================
\section{Three preliminary findings}\label{sec:findings}
% =====================================================================
We frame the empirical work as three questions. All use Qwen-2.5-3B (base, fp16)
and only the model's own logits; no training.

\paragraph{(1) Does the model model its decoder? Yes.}
We constrain the decoder while the model generates, then ask whether the model,
reading back its own constrained text, has inferred the constraint.
\textbf{Blacklist:} ban a token during generation;
the model lowers its predicted probability on that token across the sequence
($\Delta p_{\text{late--early}} = \blRankOneDeltaP$ for the top token) and below
an unbanned control ($\blRankOneTvc$, 95\% CI
$[\blRankOneTvcLo,\blRankOneTvcHi]$), with an early-position contrast
$\approx 0$ as a sanity check. \textbf{Top-$k$:} truncating to the top $k$
produces no significant cliff at $k$ itself; instead probability concentrates on
the head, the more so the tighter the cutoff. So the model does track its
decoder, though the top-$k$ signal is broad sharpening rather than a $k$-specific
edge --- a conservative reading the controls do not let us strengthen to ``infers
the exact constraint.''

\paragraph{(2) Can we use that to predict better? Yes on temperatured text, no in
the wild.} A streaming filter reads $\hat\beta_t$ off the prefix one token at a
time (Appendix~\ref{app:laplace}). On Qwen self-text generated at a \emph{known}
temperature ($\scNseq$ sequences per $T$), $\hat\beta_t$ settles on the true
$\log(1/T)$ from the prefix alone (bias $\scFaithBiasZeroSeven$ at $T{=}0.7$), and
using it as a per-token temperature beats the default $\beta{=}1$ model by
$\scNllGainZeroSeven$ nats at $T{=}0.7$ (up to $\scNllGainTwo$ at $T{=}2$), within
$\scOracleGapTwo$ nats of an oracle that \emph{knows} $1/T$. But does
\emph{natural} text carry a recoverable
temperature? Fitting the best global temperature per text type over
$\scSurvNDatasets$ types from in-distribution to far-OOD
(appendix Table~\ref{tab:survey}), Qwen is temperature-robust everywhere: optimal
$T\in[\scSurvOptTRecaman,\scSurvMaxOptT]$ and the best temperature buys
$\le\scSurvGainRecaman$ nats on any real type ($\scSurvGainShuffledWikitext$ only
on shuffled gibberish). So the read-out is real and usable on temperatured text,
but in the wild there is essentially nothing to recalibrate.

\paragraph{(3) Can we use it to generate better? Not yet.}
The tempting move is to feed $\hat\beta_t$ back into the sampler to ``correct''
the temperature as it generates. It self-neutralizes: correcting toward
$\hat\beta$ drives the effective decoding temperature back to
$\beta_{\mathrm{dec}}\to 1$, quietly cancelling the very signal it reads off. We
confirmed this across exact-grid filters, so it is not an approximation artifact.
Turning the read-out into better generation is the honest open problem.

% =====================================================================
\section{Limitations and next steps}
% =====================================================================
These are preliminary steps, on a single model and tokenizer; the certificate
certifies the effective $\beta$ of the \emph{observed} stream, never a
counterparty's dial. The natural next step, for which it is the right tool, is
misspecification: the \emph{effective} temperature and goodness-of-fit that
top-$k$ and blacklist decoders register, where no single temperature reproduces
the stream and the best-fit bucket is a projection (synthetic pilot:
\texttt{scripts/e1\_synthetic\_pilot.py}).

\appendix
% =====================================================================
\section{The fast point-estimate filter}\label{app:laplace}
% =====================================================================
The bucket certificate is a mixture forecaster; for a single-number read-out we
use a companion streaming \emph{Laplace} filter --- a prior-regularized one-step
online MLE on $\eta=\log\beta$. With a log-normal prior (mean $1$, scale
$\sigma_0$) and a per-token Newton/Fisher-scoring update of the categorical
likelihood under $q=\mathrm{softmax}(\ell_t)$, the update uses
$\text{score}_t = \ell_t[x_t]-\mathbb{E}_q[\ell_t]$ and
$J_t = \sigma_0^{-2}+\sum_{\tau\le t}\mathrm{Var}_q[\ell_\tau]$, with $q$
evaluated at $\beta=1$. This estimator is fast and is what produces the
$\hat\beta_t$ used in finding~(2); it carries \emph{no} per-sequence guarantee
(its experts are not fixed), which is exactly why the certificate of
Section~\ref{sec:cert} exists and why we validate the filter \emph{against} it
(Table~\ref{tab:grid_certificate}). Implementation and unit tests (shift
invariance, batched\,$=$\,serial, recovery under matched-prior synthetic data)
are in \texttt{src/decoding\_decoding/counter\_decode.py}.

% =====================================================================
\section{Corrector and supporting control experiments}
% =====================================================================
A corrector $\beta^{\mathrm{dec}}_t = \beta^{\mathrm{target}}/\hat\beta_t^{\,p}$
divides a user's target by the estimate. An empirical sweep shows it pins the
effective temperature in the ``boredom'' regime ($\beta^{\mathrm{target}}>1$) but
overshoots in the ``confusion'' regime ($\beta^{\mathrm{target}}<1$), where
low-$\beta$ self-generation degrades into incoherent text within $\sim$20 tokens
(a garbage-in/garbage-out confound that limits what the confusion-regime numbers
can claim). Two further controls sharpen the picture and are reported in full in
the longer draft (\texttt{paper/decoding\_paper.tex}): a sampler-switch
experiment indicating the belief is not a fast per-token entropy read-out (though
entropy and rank statistics are closely related, so this is a null on one
specific mechanism, not a positive identification), and a closed-loop-vs-%
invariance comparison whose only firmly supported conclusion is ``do not correct
at $\beta^{\mathrm{target}}=1$, because Qwen is already calibrated there.''

\begin{table}[h]
\centering\small
\input{\tabdir/self_calibration_survey.tex}
\caption{Finding (2), natural text (full table): the best global temperature buys
almost nothing over $\beta{=}1$ across $\scSurvNDatasets$ text types.
${}^{**}$ marks a gain significant at the stated level; magnitudes are negligible
on all real types.}
\label{tab:survey}
\end{table}

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
