\section{Future Work}

Several directions remain unexplored.

\paragraph{Prompt format optimization.} We use a fixed prompt template throughout and do not attempt to optimize it for ICL elicitation.
Prompt engineering could improve the metric's discriminative power.
Dimension-specific framings (e.g., ``Here are several stories in different genres'') are one such variant: priming $\theta$ to expect variation along a dimension removes the small portion of the surprise attributable to the mere existence of that variation, but it does not isolate per-dimension diversity, since the rest of the cross-response surprise reduction remains.

\paragraph{Ensembling base models.} A single base model $\theta$ may produce non-monotone $a_k$ curves when pushed out of distribution by long contexts.
Ensembling multiple base models at the token level (averaging softmax probabilities) could stabilize the estimates while preserving the autoregressive structure.
Our codebase supports token-level ensembling, but we have not yet validated it experimentally.

\paragraph{Disagreement cases with other metrics.}
We report aggregate correlations with embedding-based and reference metrics on Tevet's tasks (Section~\ref{sec:tevet}) but do not look at \emph{which} response sets the metrics rank differently.
Identifying prompts or response sets where $C \times a_n$ disagrees with SentBERT, distinct-$n$, or McDiv is the most direct way to characterise what each metric is actually measuring; the disagreements are where the choice of metric matters.
We expect two regimes to drive disagreement: response sets that share arbitrary rare patterns a base model can pick up on but that fall outside what a fixed sentence embedder represents (where $C \times a_n$ should detect repetition that embeddings miss), and response sets where one response is a mixture of two others (where progressive conditioning credits the mixture for being predictable from its components, while pairwise embedding-distance averages do not).

\paragraph{Total-bits variants of $D$.}
We report $D_{Ca_n} = C \times a_n$ in bits/byte for length control, normalising by bytes rather than tokens so that scores remain tokenizer-agnostic.
The implementation also exposes $a_n$ in total bits, which is likewise tokenizer-agnostic but not length-controlled.
A hybrid scalar $C \times a_n^{\text{bits}}$ rewards length (longer responses have more positions at which to be surprising), which some practitioners may want; characterising when each variant is the appropriate measurement target is left to future work.

\paragraph{Optimizing generators against $D$.}
We use $C \times a_n$ as an evaluation-time diagnostic, but it could in principle serve as a training or decoding objective. Future work could fine-tune or sample from a generator $\pi$ with the goal of maximizing $C \times a_n$ under a fixed base model $\theta$ to push $\pi$ toward response sets that are both coherent and diverse.

