\section{Scaling the Base Model: Qwen2.5-3B vs Qwen3-30B-A3B-Base}\label{app:qwen3-comparison}

To test whether a stronger base model improves the metric on Tevet's benchmarks, we re-ran the full evaluation using Qwen3-30B-A3B-Base \citep{yang2025qwen3} (a 30B-parameter mixture-of-experts model with $\sim$3B active parameters per token) via the Tinker API,\footnote{\url{https://thinkingmachines.ai/tinker}} with the same setup as in Section~\ref{sec:tevet}: completion-format prompting, 50 permutations, no fine-tuning.
Table~\ref{tab:qwen3-comparison} reports per-task $C \times a_n$ (per-byte) for both base models.

\begin{table*}[!t]
\centering
\caption{Per-task comparison of $C \times a_n$ (per-byte) for Qwen2.5-3B vs Qwen3-30B-A3B-Base on Tevet diversity-eval. $\Delta$AUC is positive when the larger model wins.
ROC AUC is reported for the binary tasks (McDiv, McDiv\_nuggets, ConTest); only Spearman $\rho$ is meaningful for DecTest (continuous temperature labels).}
\label{tab:qwen3-comparison}
\small
\input{results/tables/qwen3_comparison.tex}
\end{table*}

\paragraph{Result: scaling did not help on the binary tasks.} On the \qwenThreeBinaryTotal{} binary classification tasks (McDiv, McDiv\_nuggets, ConTest), Qwen2.5-3B wins on \qwenThreeBinaryWinsQwenTwoFive{} of \qwenThreeBinaryTotal{}, with ConTest prompt\_gen the only marginal win for Qwen3-30B-A3B ($\qwenThreeConTestPromptGenDeltaAUC$ AUC).
The mean degradation from scaling is small but consistent (mean $\Delta$AUC = $\qwenThreeBinaryMeanDeltaAUC$ across the \qwenThreeBinaryTotal{} binary tasks).
On the \qwenThreeDecTestTotal{} DecTest tasks (continuous temperature labels), Qwen3-30B-A3B is mildly better on \qwenThreeDecTestWins{} of \qwenThreeDecTestTotal{} (mean $\qwenThreeDecTestMeanDeltaRho$ in Spearman $\rho$, with one tie).

\paragraph{Interpretation.} The binary result is a clear negative: scaling from 3B to 30B active parameters does not improve discrimination on McDiv or ConTest.
We do not have a confident explanation.
The Qwen3-30B-A3B run was not repeated after a bug fix applied to the primary experiments, so the comparison should be treated as preliminary.
We leave the question of whether larger base models improve the metric on binary tasks to future work.
