\section{Method}\label{sec:method}

This section defines the three quantities used in the main body: the progressive conditional surprise curve $a_k$, the coherence term $C$, and the scalar score $D_{Ca_n} = C \times a_n$.
The full information-theoretic motivation, the alternative excess-entropy summary $E$, and the cross-mode learning analysis appear in Appendices~\ref{app:excess-entropy}--\ref{app:qwen3-comparison}.

\subsection{Progressive Conditional Surprise}\label{sec:progressive-conditioning}

Given prompt $p$ and $n$ responses $r_1, \ldots, r_n$ from policy $\pi$, define
\begin{equation}\label{eq:ak}
    a_k = -\log_2 \theta(r_k \mid r_{<k}, p) \quad \text{for } k = 1, \ldots, n,
\end{equation}
the total surprise of the $k$-th response under base model $\theta$ given the previous $k{-}1$ responses.
The sequence $(a_1, \ldots, a_n)$ is the \textbf{progressive conditional surprise curve}.
We normalize each $a_k$ by the byte count $\|r_k\|$ to get a per-byte (bits/byte) curve that is independent of $\theta$'s tokenizer.
All quantities below are per-byte unless otherwise stated.

The intuition is that as $\theta$ sees more responses from $\pi$, the conditional surprise drops if the responses share patterns ($\theta$'s in-context learning picks up on modes, stylistic regularities, topic patterns) and stays flat if they do not.
For a policy with rich coherent diversity the curve declines and levels off at a positive floor; for a policy producing one repeated mode the curve drops sharply to near zero; for pure noise the curve stays roughly constant at a high value.

In practice we average $a_k$ over uniformly random permutations of the response ordering (so each position averages over all responses and the curve reflects only how $\theta$'s predictions improve with more context, not which response happened to land in slot $k$).
We take the metric value to be the last observed point $a_n$.
Equivalently, $a_n = H_\theta(r_n \mid p) - I_\theta(r_n; r_1, \ldots, r_{n-1} \mid p)$: the response's individual surprise under $\theta$ minus how much the prior responses make it predictable.
High $a_n$ therefore requires $r_n$ to be both individually surprising and not made predictable by the others, two properties one would want a diversity score to reward.
No fitting or extrapolation step is involved.
The curve $a_k$ is itself informative: its shape distinguishes a one-mode collapse (sharp drop to a low floor) from richer diversity (gradual decline to a higher floor) in ways the endpoint alone obscures.

\subsection{The Coherence Term}\label{sec:coherence}

The curve alone is fooled by pure noise: $\theta$ can never predict random tokens from one another, so $a_n$ stays at the high unconditional surprise level and noise would dominate the metric.
The distinguishing signal lies in how plausible $\theta$ finds each individual response, independent of the others.

Let $h_\theta(r \mid p) = -\frac{1}{\|r\|}\log_2 \theta(r \mid p)$ be the per-byte cross-entropy of response $r$ under $\theta$ conditioned on the prompt alone.
Define \textbf{coherence} as the geometric mean of the per-byte probabilities:
\begin{equation}\label{eq:coherence}
    C = 2^{-\frac{1}{n}\sum_{i=1}^{n} h_\theta(r_i \mid p)} = \frac{1}{\mathrm{PPL}_\theta(\pi, p)},
\end{equation}
the reciprocal of the geometric-mean per-byte perplexity\footnote{In our Tevet evaluation with Qwen2.5-3B, the central 90\% (5th to 95th percentile) of per-response bits-per-byte values across \tevetCoherenceRespCount{} responses span $[\tevetCoherenceBpbLow, \tevetCoherenceBpbHigh]$, giving $C \in [\tevetCoherenceCLow, \tevetCoherenceCHigh]$; the mean per-set $C$ across \tevetCoherenceSetCount{} response sets is \tevetCoherenceMeanC.} that $\theta$ assigns to the responses individually.\footnote{For scale, the bits/byte score is the average per-byte cross-entropy $\bar{\ell}$, related to coherence by $C = 2^{-\bar{\ell}}$. GPT-3 davinci on The Pile \citep{gao2020pile} scored $\sim$0.72 bits/byte and GPT-2 XL $\sim$1.05 bits/byte, giving $C \approx 0.61$ and $C \approx 0.49$ respectively; stronger base models reach lower bits/byte and correspondingly higher $C$.}
Perplexity is a standard metric of incoherence, and the geometric form is intended to suppress sets containing incoherent responses: a single sample with high per-byte cross-entropy drives $C$ toward zero, limiting the rescue effect of any single fluent response on an otherwise incoherent set.

\subsection{The Diversity Score}\label{sec:scalar}

The scalar score is the product:
\begin{equation}\label{eq:D-Can}
    D_{Ca_n} = C \times a_n \quad \text{(bits/byte).}
\end{equation}
It can be read as ``how many bits of surprise per byte remain after $\theta$ has learned what it can from the other responses, weighted by output plausibility.''
The intended edge-case behaviors are summarized in Table~\ref{tab:edge-cases}, alongside empirical $D_{Ca_n}$ on synthetic instances of each scenario under two base models (full per-metric breakdown in Appendix~\ref{sec:scenario-validation}, Table~\ref{tab:scenarios}).

\begin{table*}[t]
\centering
\small
\caption{Intended edge-case behavior of $D_{Ca_n} = C \times a_n$ on the five synthetic scenarios (Appendix~\ref{sec:scenario-validation}), with empirical $D_{Ca_n}$ values under GPT-2 (124M) and Qwen2.5-3B base.
The product correctly suppresses pure noise and incoherent multi-mode (via $C$) and one-mode sets (via $a_n$); multi-mode coherent is the intended winner.
Empirical numbers show smaller $a_n$ for multi-mode coherent than this story predicts: with only 3 distinct modes in a set of 10 responses, $\theta$'s in-context learning identifies them and $a_n$ is no longer high. In such ambiguous cases reporting the full $a_k$ curve is more principled than the scalar $a_n$: a different choice of $n$ would reveal the diversity, and showing the curve sidesteps having to pick that $n$ in advance. Appendix~\ref{sec:mode-count-scaling} shows $a_n$ rising with mode count.
Mixed empirically scores highest (via a high $a_n$) rather than the predicted ``mid'' position; reweighting strategies are discussed in Section~\ref{sec:exp-discussion}. On Qwen2.5-3B, multi-mode incoherent also outranks multi-mode coherent; Section~\ref{sec:scalar} discusses.}
\label{tab:edge-cases}
\begin{tabular}{@{}lccccc@{}}
\toprule
                      & \multicolumn{3}{c}{Predicted}            & \multicolumn{2}{c}{Empirical $D_{Ca_n}$} \\
\cmidrule(lr){2-4} \cmidrule(lr){5-6}
Scenario              & $C$  & $a_n$ & $D_{Ca_n}$              & GPT-2  & Qwen2.5-3B \\
\midrule
Pure noise            & low  & high  & low (via $C$)           & \scenarioPureNoiseGptDCan{}        & \scenarioPureNoiseQwenDCan{}        \\
Multi-mode incoherent & low  & high  & low (via $C$)           & \scenarioMultiIncoherentGptDCan{}  & \scenarioMultiIncoherentQwenDCan{}  \\
Multi-mode coherent   & high & high  & \textbf{high}           & \scenarioMultiModeGptDCan{}        & \scenarioMultiModeQwenDCan{}        \\
One-mode              & high & low   & low (via $a_n$)         & \scenarioOneModeGptDCan{}          & \scenarioOneModeQwenDCan{}          \\
Mixed                 & mid  & high  & mid                     & \scenarioMixedGptDCan{}            & \scenarioMixedQwenDCan{}            \\
\bottomrule
\end{tabular}
\end{table*}

The score depends on $p$, $\pi$'s responses, and $\theta$.
Diversity is therefore a property of (responses, prompt, scoring model), not of the policy in isolation: the same response set can score differently under a stronger or weaker $\theta$.
This is by design: $\theta$'s in-context learning capability is the lens through which diversity is measured, giving the score the potential to improve as base models improve (see Appendix~\ref{app:qwen3-comparison} for a preliminary scaling experiment).
For example, Table~\ref{tab:edge-cases} shows Qwen2.5-3B scoring multi-mode incoherent (\scenarioMultiIncoherentQwenDCan{}) above multi-mode coherent (\scenarioMultiModeQwenDCan{}), reversing our predicted ranking; Figure~\ref{fig:scenario-curves} shows Qwen's $\bar{a}_k$ curve dropping over $k$ on the multi-mode incoherent scenario, indicating that its in-context learning recognises the shared template structure as a pattern despite the within-response scrambling.
GPT-2's weaker in-context learning fits the predicted ordering on this row better (\scenarioMultiIncoherentGptDCan{} for incoherent vs.\ \scenarioMultiModeGptDCan{} for coherent), but we still recommend stronger base models when possible -- sophisticated patterns are exactly what an ICL-based metric should be able to detect.
When comparing policies across a prompt suite, $D_{Ca_n}$ can be averaged over prompts or differenced; see Appendix~\ref{app:aggregation}.

