\section{Practical Considerations}\label{sec:practical}

\subsection{Formatting the Conditioning Context}\label{sec:formatting}

Computing $\theta(r_k \mid r_{<k}, p)$ requires feeding $\theta$ a context containing the prompt and previous responses.
The formatting choice affects results, and we use two formats depending on whether the responses are best read as parallel answers to the prompt or as continuations of it.
Both are implemented as \texttt{format\_conditioning\_context} in \texttt{src/icl\_diversity/core.py} and selected via a \texttt{format\_mode} argument throughout the pipeline.

\paragraph{Instruct format.} The default, used for the synthetic scenarios (Appendix~\ref{sec:scenario-validation}), the mode-count experiments (Appendix~\ref{sec:mode-count-scaling}), and the OLMo-2-7B post-training case study (Section~\ref{sec:rlhf-experiment}):

\begin{verbatim}
    [prompt p]

    Response A: [r_1]

    Response B: [r_2]

    Response C: [r_3]
\end{verbatim}

The prompt appears once and each response is introduced by a labelled header (``Response A:'', ``Response B:'', \ldots, with labels rolling over to ``AA'', ``AB'' past ``Z''), encouraging the base model's in-context learning to treat the responses as parallel answers to the same prompt.

\paragraph{Completion format.} Used for the Tevet and Berant evaluation (Section~\ref{sec:tevet}), where each response set is a continuation of a shared narrative prompt rather than an instruction-style answer:

\begin{verbatim}
    1. [prompt p] [r_1]

    2. [prompt p] [r_2]

    3. [prompt p] [r_3]
\end{verbatim}

The prompt is repeated immediately before each response so $\theta$ scores $r_k$ in the same prompt-as-context that produced it.
Per-token log-probabilities are extracted only over the response tokens; the bits accumulated over the repeated prompt prefix are excluded from $a_k$, keeping the bits-numerator and the per-byte denominator both attributable to $r_k$ alone.

\paragraph{Base vs.\ instruction-tuned models.} We use a base (non-instruction-tuned) model as $\theta$ for two reasons.
First, an instruction-tuned model has had its output distribution shaped by RLHF or similar procedures, so using one as $\theta$ reintroduces the kind of distributional bias we seek to avoid.
Second, an instruction-tuned $\theta$ may assign systematically different probabilities to text that follows or violates its alignment training, introducing a confound between coherence-as-fluency and coherence-as-alignment that the metric should not conflate.
Modern base models already exhibit strong in-context learning from pretraining alone, so this choice does not sacrifice ICL capability.
Optimizing the prompt format to maximally elicit in-context learning from $\theta$ is an interesting direction but out of scope for this paper; we use the neutral ``Response A/B/C'' template throughout.

\subsection{Computational Cost}\label{sec:cost}

Because $\theta$ is a causal language model, the full $a_k$ curve for one response ordering can be computed from a \textbf{single forward pass}.
The prompt and all responses are concatenated into one sequence (Figure~\ref{fig:single-pass-layout}).
The forward pass produces the log-probability of every token conditioned on all preceding tokens.

\begin{figure*}[t]
\centering
\begin{verbatim}
[prompt p] Response A: [r_1] Response B: [r_2] ... Response N: [r_n]
\end{verbatim}
\caption{Single-pass input layout for computing the full $a_k$ curve.
The prompt is followed by all $n$ responses with ``Response A/B/C/\ldots'' labels; one forward pass over this sequence yields every $a_k$ value simultaneously, since causal attention conditions the tokens of $r_k$ on exactly $p, r_1, \ldots, r_{k-1}$ (plus formatting).}
\label{fig:single-pass-layout}
\end{figure*}
The tokens belonging to $r_k$ are conditioned on exactly $p, r_1, \ldots, r_{k-1}$ (plus formatting), which is the conditioning required for $a_k$.
Partitioning the output log-probabilities by response boundary and summing within each partition yields all $n$ values of $a_k$ simultaneously for that ordering, with FLOPs scaling as $O((|p| + n\bar{L}_{\mathrm{tok}})^2)$ from causal attention, where $\bar{L}_{\mathrm{tok}}$ is the average response length in tokens.
Because we report $\bar{a}_k$ averaged over $|\Sigma|$ random permutations of the response ordering (Section~\ref{sec:ordering}), the total $a_k$-curve cost is $|\Sigma|$ such passes; the permutations are independent and we batch them on a single GPU.

The coherence term $C$ additionally requires the \emph{unconditional} per-byte cross-entropies $h_\theta(r_i \mid p)$ for each response.
These cannot be extracted from the concatenated pass, since in that pass $r_i$ for $i > 1$ is conditioned on all prior responses.
Computing them requires $n$ independent forward passes, each over the short context $(p, r_i)$.
These passes are embarrassingly parallel and batchable, and they do not depend on the ordering, so $C$ is computed once per response set regardless of $|\Sigma|$.

In summary, scoring one (prompt, response set) tuple takes:
\begin{itemize}[nosep]
    \item $a_k$ curve: $|\Sigma|$ forward passes (long context, one per permutation).
    \item $C$: $n$ forward passes (short contexts, batchable; not multiplied by $|\Sigma|$).
\end{itemize}
With $n = 10$ responses and $|\Sigma| = 50$ permutations (our default for larger-scale experiments), this is $50 + 10 = 60$ forward passes per (prompt, response set) tuple; the long-context $|\Sigma|$ passes dominate wall-clock time.

\subsection{Dependence on Sample Ordering}\label{sec:ordering}

The $a_k$ values depend on the ordering of $\{r_i\}$.
Individual responses differ in how surprising they are to $\theta$ given the responses that precede them, so $a_k$ depends on which response sits at position $k$ and which others precede it, making a single ordering's curve jagged.
Averaging over random permutations removes this dependence: each position averages over many choices of response and preceding context, so the curve reflects only how $\theta$'s predictions improve with more context.
We therefore average over $|\Sigma|$ random permutations; Section~\ref{sec:practical-findings} compares low- and high-permutation runs on the validation scenarios and finds that the low setting misranks adjacent scenarios.
We have not swept intermediate values and cannot pinpoint a cheap-and-reliable minimum, so we default to $|\Sigma| = 100$ for scenario-level experiments and $|\Sigma| = 50$ elsewhere.

Per-response normalization by byte count must be performed \emph{before} averaging across permutations, since the byte count depends on which response lands at each position.
Concretely, the per-byte curve is a mean of per-permutation rates,
\begin{equation}\label{eq:akbar-mor}
    \bar{a}_k = \frac{1}{|\Sigma|}\sum_{\sigma \in \Sigma} \frac{a_k^{\sigma}}{\|r_{\sigma(k)}\|},
\end{equation}
not a ratio of permutation-averaged bits to permutation-averaged byte counts (which would degenerate, after enough permutations, into the total-bits curve rescaled by the mean response length).

\subsection{Choice of $n$}

The diversity score depends on $n$ as a measurement parameter, not through any estimate of an asymptotic floor.
Given enough in-context examples, even genuinely diverse policies eventually become predictable to $\theta$, so $a_k$ keeps decreasing rather than converging to a meaningful irreducible value: there is no $a_\infty$ for $a_n$ to approximate.
Too small an $n$ means $\theta$ has not yet exploited the inter-response structure, and $D_{Ca_n}$ overstates diversity.
Too large an $n$ means $\theta$ has accumulated enough in-context examples to reduce its surprise even within genuine diversity; for any policy with finitely many distinct modes, $a_n$ continues to decrease toward zero as $\theta$ learns to predict within-mode variation from prior examples.
For comparisons to be meaningful, $n$ must be held fixed across all policies under evaluation.
Our experiments use $n = 5$ on Tevet (Section~\ref{sec:tevet}), $n = 10$ on OLMo-2-7B and scenario validation, and $n = 20$ on the mode-count experiment (Appendix~\ref{sec:mode-count-scaling}).
This is one reason to report the full $a_k$ curve rather than $D_{Ca_n}$ alone. $D$ is a lossy summary, but the curve shows directly how response surprise evolves at every context size.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SECTION 8: EXPERIMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
