\section{Tevet--Berant Diversity-Eval: Human-Grounded Validation}\label{sec:tevet}

A diversity metric should be tested against human judgements on a standardized eval; \citet{tevet2020evaluating} provide both.
Their McDiv and ConTest datasets pair human-grounded diversity labels with a fixed comparison protocol (Spearman $\rho$ and OCA: \emph{optimal classification accuracy}, the best accuracy achievable by a one-dimensional threshold separating low- and high-diversity sets), the standard methodology for ranking diversity metrics against human judgements.
The response sets are written by Mechanical Turk workers, and for both McDiv\_nuggets and ConTest the high/low label is fixed by the construction protocol itself: a worker writes five different continuations (the high-diversity set), then self-selects one of those continuations and paraphrases it five times preserving content but varying form (the low-diversity set; see Appendix~\ref{app:confound} for the full protocol and a construction-confound caveat).
We run $D_{Ca_n}$ through this evaluation pipeline.

\paragraph{Setup.} We use Qwen2.5-3B (base) with 50 permutations in completion format (Appendix~\ref{sec:formatting}) on the released splits: \tevetMcDivSetCount{} McDiv sets (no\_hds), \tevetMcDivNugSetCount{} McDiv\_nuggets sets (no\_hds), and \tevetConTestSetCount{} ConTest sets (with\_hds), each with 5 worker-written responses per set and a binary high/low diversity label.
We follow Tevet and Berant in reporting Spearman $\rho$ and OCA, and additionally report ROC AUC (the natural metric for binary classification).

\begin{table*}[t]
\centering
\caption{ConTest results: Spearman $\rho$ and OCA between a set's diversity class and each metric score.
Baselines from Tevet's pre-computed metrics; $C \!\times\! a_n$ and $a_n$ are ours (Qwen2.5-3B, 50 permutations, per-byte).
Best result per task in bold.}
\label{tab:contest}
\small
\input{../results/tables/contest_rho_oca.tex}
\end{table*}

\paragraph{Headline result.} On the binary tasks $D_{Ca_n}$ clearly tracks diversity, with OCA between \tevetCxAnVsSentBertOCAMinPct\% and \tevetCxAnVsSentBertOCAMaxPct\% behind SentBERT (Table~\ref{tab:contest}).
On McDiv prompt\_gen, where $D_{Ca_n}$ performs best, it reaches $\rho = \tevetMcDivPromptGenCxAnRho{}$ and OCA = \tevetMcDivPromptGenCxAnOCA{}, against SentBERT's $\rho = \tevetMcDivPromptGenSentBertRho{}$ and OCA = \tevetMcDivPromptGenSentBertOCA{}; the corresponding ROC AUC is \tevetMcDivPromptGenAUC{}.
The headline takeaway is that an ICL-based diversity metric reaches close to a sentence-embedding baseline on a human-grounded eval, with no embedding model, no reference corpus, and no human labels in the metric itself.

\paragraph{DecTest, included for completeness.} Tevet and Berant also release DecTest (Table~\ref{tab:dectest}), a diagnostic where the ``label'' is the sampling temperature used to generate each response set rather than a human judgement.
We include it for parity with the released benchmark only: we do not believe that temperature labels map onto creative/meaningful output diversity, consistent with \citet{peeperkorn2024temperature}'s finding that the relationship between sampling temperature and creativity is more nuanced and weaker than the ``creativity parameter'' framing suggests.
The raw $a_n$ achieves $\rho = \tevetDecTestAnPromptGenRho{}$ on prompt\_gen and $\rho = \tevetDecTestAnRespGenRho{}$ on resp\_gen, matching or exceeding all baselines.

\begin{table}[t]
\centering
\caption{DecTest results: Spearman $\rho$ between sampling temperature and each metric score (1000 samples, no\_hds).
Reported for parity with Tevet and Berant; the labels here are sampling temperatures rather than human judgements, so DecTest is not part of the framing for $D_{Ca_n}$.
That $a_n$ tracks temperature is mechanically unsurprising: $a_n$ is a conditional-entropy quantity, and sampling temperature directly scales the entropy of the policy, so a positive $\rho$ here is a sanity check rather than evidence about creative diversity; \citet{peeperkorn2024temperature} likewise find only a weak relationship between temperature and creativity in LLM outputs.}
\label{tab:dectest}
\small
\input{../results/tables/dectest_rho.tex}
\end{table}

\paragraph{Caveats.} The McDiv\_nuggets construction protocol introduces a confound: low-diversity sets are paraphrases of specific, dramatic endings that are intrinsically more surprising to the base model than typical high-diversity continuations.
Appendix~\ref{app:confound} documents the mechanism, the per-byte $a_1$ gap, and an evidence-by-length-bin breakdown showing the effect is content-driven not length-driven.
The $C$ weighting in $D_{Ca_n}$ helps on this benchmark partly for that confound-related reason; on settings without such a confound we expect $a_n$ to carry most of the signal.

