\section{Limitations}\label{sec:limitations}

We collect the main limitations of the framework and of our $D = C \times a_n$ choice in one place so readers can calibrate how strongly to update on the results above.
Several points below recap caveats already made in their natural context and cross-reference those for detail.

\paragraph{Measurement is relative to $\theta$'s perception.}
The metric evaluates diversity \emph{as perceived by the base model $\theta$}: if $\pi$'s outputs differ only in ways $\theta$ cannot distinguish from context, the metric underestimates diversity (Section~\ref{sec:motivation}, ``Important caveat'').
Concretely, the off-diagonal sign of the pairwise surprise-reduction matrix changes with $\theta$'s scale (Section~\ref{sec:cross-mode-learning}), so $a_k$ curves for the same response set are not directly comparable across different choices of $\theta$.

\paragraph{Metric selection saw the full Tevet benchmark.}
We chose $C \times a_n$ (rather than alternative summaries such as $C \times E$; Appendix~\ref{app:excess-entropy}) after observing its performance on the full Tevet diversity-eval benchmark, without holding out a validation split.
The Tevet numbers in Section~\ref{sec:tevet} may therefore be optimistically biased: the scalar form was selected with visibility into those results.
The OLMo-2-7B post-training experiment (Section~\ref{sec:rlhf-experiment}) was scored after the metric form was fixed and provides independent validation; the synthetic mode-count experiments (Section~\ref{sec:mode-count-scaling}) informally influenced the choice (had they failed, we would likely have revised the metric) and so are not strictly independent.

\paragraph{External benchmark performance trails the strongest baseline.}
On Tevet and Berant's diversity-eval benchmark, $C \times a_n$ trails SentBERT by \tevetCxAnVsSentBertOCAMinPct\%--\tevetCxAnVsSentBertOCAMaxPct\% OCA across the nine binary tasks (Section~\ref{sec:tevet}); scaling the base model to Qwen3-30B-A3B-Base \citep{yang2025qwen3} did not close the gap (Appendix~\ref{app:qwen3-comparison}).
We position ICL-based diversity metrics (of which $D_{Ca_n}$ is one instance we found to work well) as a principled complement to embedding-based metrics, not a replacement.

\paragraph{Length-matching drops short-response prompts.}
Reporting in bits/byte (rather than total bits) makes per-byte cross-entropy not invariant to response length, so the OLMo-2-7B experiment scores a length-matched subset and drops $111/300$ prompts whose common per-prompt byte length falls below $50$~bytes; full mechanics and the un-truncated robustness check are in Section~\ref{sec:exp-discussion}.

\paragraph{$D = C \times a_n$ is a pragmatic choice, not a theoretically derived one.}
The product form was selected because it captures the properties we wanted a diversity score to have (residual surprise remains after the base model has seen previous responses, low-coherence outputs are penalised, and the result stays on a bits-per-byte scale), not because it is derived from an axiomatic information-theoretic setup.
Other scalars respecting the same constraints (weighted integrals of $a_k - a_\infty$, slope-based scores, or coherence applied at a different point on the curve) could equally be defended (see ``The framework admits other metrics'' in Section~\ref{sec:exp-discussion}).
We adopt $D_{Ca_n}$ because it is interpretable and works empirically across our validation regimes, not because it is uniquely correct.
We invite other researchers to build on this work to develop other formulas that characterize the $a_k$ curve in meaningful ways.